Operational constraints and closing civic space: International NGO restrictions and China’s Overseas NGO law

### 2023-12-05 manuscript/manuscript.qmd
Shrink/edit intro and legal framework
Word count is now under 9,000!

In the past two decades, nearly 100 countries have proposed or enacted legal measures designed to restrict nongovernmental organizations (NGOs) and civil society more broadly [@Chaudhry:2022; @ChaudhryHeiss:2021; @CarothersBrechenmacher:2014; @DupuyPrakash:2018; @DupuyPrakash:2020; @DupuyRonPrakash:2016].[^ngo-clarification] According to the CIVICUS Civil Society Monitor, as of 2023 only `r civicus_prop$true$prop_nice` of countries have open and unrestricted civic space [@ChaudhryHeiss:2022a; @CIVICUSMonitor]. This global trend of closing or shrinking civic space has affected countries of all regime types, including both autocracies and democracies such as Morocco and Palestine [@AtiaHerrold:2018], India [@press-trust-india:2019], Vietnam [@Sidel:2023], Canada [@CIVICUS:2013], and the United States [@Cramer:2020]. On April 28, 2016, China passed its own set of civil society restrictions with its Law of the People's Republic of China on Administration of Activities of Overseas Nongovernmental Organizations in the Mainland of China---or "2016 ONGO Law"---which formally took effect on January 1, 2017. The law requires all international NGOs (INGOs) to either establish formal in-country representative offices (ROs) or file short-term temporary activities to lawfully operate in mainland China. Without recognized legal status, INGOs are prohibited from working in the country, and even with legal recognition, INGO programming is limited to government-approved activities.

[^ngo-clarification]: "Civil society" is a broad concept that includes a range of organizational forms, both formal and informal. In this paper, we look at nongovernmental organizations (NGOs), often called nonprofit organizations or third sector organizations---formal non-profit-seeking organizations that are neither privately-owned nor government-controlled. We define international NGOs as organizations that conduct work predominantly in at least one other country outside of their home country.

As an authoritarian state with a growing nonprofit sector, China's regulatory regime for INGOs serves as an important example for understanding and exploring how autocracies use restrictive laws as policy tools as they engage with international civil society. The restrictive policies implemented by China's 2016 ONGO law serve dueling purposes that both harness the benefits and reduce the risks of INGOs. The law is designed to decrease demand for foreign civil society organizations and limit the influence they might have over domestic policies [@DeMattee:2018; @DeMattee:2019], while also allowing the regime to benefit from the expertise, services, and reputational advantages of working partnering with international civil society [@Plantan:2022; @Heiss:2019]. These dual demands lead to differential enforcement of civil society laws and a divergence between formal *de jure* regulations and the *de facto* implementation of those laws [@ChaudhryHeiss:2022], as the government enforces civil society regulations selectively depending on regime preferences for INGO services.

In this paper, we use the case of China to explore how the dual demands to both restrict and embrace INGOs have shaped the international nonprofit sector in the five years since the 2016 ONGO Law came into effect. We examine how the ONGO Law has influenced the population of INGOs that have successfully registered in China, as well as the association between factors such as INGO issue areas and the operational space granted to them in the country. Through our study, we make several contributions to a growing body of research on the effects of closing civic space on INGO behavior and strategies. First, we expand work that uses institutional approaches to understanding civil society regulations, which argues that states strategically regulate INGOs to maintain regime stability, choosing forms of regulation and repression that minimize threats to internal political stability [@Chaudhry:2022; @DeMattee:2018; @DeMattee:2019; @Heiss:2019]. Second, we explore how authoritarian civil society regulations act as policy tools that regimes can leverage to shift INGO behavior in their favor and how restrictive laws can reshape and tame organizational strategies [@Li:2020; @WuNie:2021; @LianMurdie:2023]. Third, we explore how these regulations lead to differentiated application of the law depending on organizational characteristics. We argue that INGO issue areas, missions, and pre-existing relationships with local government officials influence the degree of operating space available for INGOs.

By focusing on INGO operational space in China, we provide a unique empirical view into the *de facto* regulatory environment for foreign civil society in an authoritarian state. Many previous studies use broad indexes of civil society openness, such as the core civil society index from the Varieties of Democracy project [@LianMurdie:2023; @ChaudhryHeiss:2022]. These measures capture the general regulatory environment for all civil society---including domestic NGOs, INGOs, and informal organizations---but do not reflect the specific implementation of laws for INGOs. In this paper, though, we use a mixed methods approach to explore the regulatory environment for INGOs in particular, providing a much richer description and analysis of the impact of INGO laws.

Below, we provide a brief overview of China's 2016 ONGO Law and the different regulatory mechanisms---and practical limitations---it provides for INGOs to obtain legal access to the country. We then summarize existing research on authoritarian regulation of INGOs and the differential application of these laws based on regime preferences and organizational characteristics, and propose a set of empirical expectations based on this literature. We test these expectations with a mixed methods research design, combining Bayeisan analysis of administrative data from `r nrow(ongo)` formally registered INGOs with a comparative case study of Greenpeace and The Nature Conservancy, two INGOs working on similar issue areas which took different approaches to operating in China after 2017. We conclude with a broader discussion of authoritarian civil society regulations and propose future methods for exploring the practical operating environment for INGOs working in such regimes.

@@

International NGOs began to play an important role in China in the early 1990s as part of the country's Reform and Opening-Up policy, and organizations such as Save the Children and the Ford Foundation established local offices. Since then, INGOs and global civil society have played pivotal roles in providing funding, organizational and program management skills, as well as catalytic support for the emergence of China's domestic nonprofit sector [@Yin:2009]. Prior to the implementation of the 2016 ONGO Law, regulation of INGOs in China was ambiguous [@Shieh:2018; @Spires:2022]. The only legal document relevant to INGOs was a section of the Regulation on the Management of Foundations that specifies the registration of ROs of foreign foundations with the Ministry of Civil Affairs, the regulator of Chinese domestic nonprofits. However, the actual application of the regulation was very limited: until the ONGO Law took effect in 2017, there were fewer than 30 foreign foundation ROs registered, while some Chinese officials estimated in 2016 that there were roughly over 7,000 INGOs operating in China [@YeHuang:2018; @Spires:2020].

Many INGOs gained legal status by registering as foreign-invested companies [@Sidel:2016], for the administrative convenience such as hiring employees [@HolbigLang:2022; @Yin:2009]. However, such a legal form is not quite compatible with the non-profit-distributing nature of INGOs. Chinese companies are intended to be for-profit only, while nonprofits are regulated under other legal forms, such as foundations and social service organizations [@Ye:2021]. Thus, INGOs registered ad foreign-invested companies had to navigate through for-profit institutional logics, with doctrines like protecting shareholders' investment returns.

The legal framework governing INGOs was overhauled by the 2016 ONGO Law. The new law applies to all overseas (including Hong Kong, Macau, and Taiwan) NGOs that conduct not-for-profit activities in the fields of economic development, education, science, technology, culture, public health, sports, environment protection, poverty alleviation, and disaster relief. The ONGO Law provides two pathways for INGOs to lawfully operate in China (see @fig-ro-ta-pathways): (1) submitting an application for formal permanent registration as ROs, or (2) filing documentation of temporary activities with the Public Security Bureaus. Formal registration as an RO allows INGOs to maintain an ongoing presence in the country, while temporary activities are intended for INGOs that need to conduct one-off programs that last no longer than one year. Using one of these two channels to obtain legal status became mandatory in 2017.

```{r fig-ro-ta-pathways, fig.width=6, fig.height=3}
#| echo: false
#| include: true
#| warning: false
#| message: false
#| fig-cap: "Two possible pathways for international NGOs to legally work in 
#|   China under the 2016 ONGO law, either as formal Representative Offices 
#|   or by undertaking specific Temporary Activities"

@@

Registering as an RO under the ONGO Law is a complex process that requires the consent of a Chinese Professional Supervisory Unit (PSU)---a governmental or quasi-governmental unit related to an INGO's issue area which oversees and co-supervises the INGO with Public Security Bureaus. Obtaining the consent of a PSU to apply for registration has become a key challenge for INGOs interested in formal registration, since potential PSUs do not have a legal obligation to assume such a demanding role [@Jia:2017; @Shieh:2018; @YeHuang:2018]. Organizations vary in their ability to attract partner PSUs, and INGOs that are unable to find suitable PSUs can end up not being able to registered or registered with great limitation in terms of work areas and geographical reach. INGOs are authorized to operate within China on a province-by-province basis. Many organizations are allowed to work in all 32 mainland provincial administrative regions, but many are also only authorized to work in a handful, or even only a single province. Similarly, INGOs need to find an eligible Chinese partner, including governmental, quasi-governmental and non-profit organizations to file temporary activities for them. After successfully gaining legal status, INGOs also need to abide by the operational regulations by the law, such as no fundraising within mainland China, only channeling the funds through RO's registered account or the Chinese partners' accounts, filing reports regularly and stay within the geographical and issue scopes as approved.

>>>>

In the past two decades, nearly 100 countries have proposed or enacted legal measures designed to restrict nongovernmental organizations (NGOs) and civil society more broadly [@Chaudhry:2022; @ChaudhryHeiss:2021; @CarothersBrechenmacher:2014; @DupuyPrakash:2018; @DupuyPrakash:2020; @DupuyRonPrakash:2016].[^ngo-clarification] According to the CIVICUS Civil Society Monitor, as of 2023 only `r civicus_prop$true$prop_nice` of countries have open and unrestricted civic space [@ChaudhryHeiss:2022a; @CIVICUSMonitor]. On April 28, 2016, China passed its own set of civil society restrictions with its Law of the People's Republic of China on Administration of Activities of Overseas Nongovernmental Organizations in the Mainland of China---or "2016 ONGO Law"---which formally took effect on January 1, 2017. The law requires all international NGOs (INGOs) to either establish formal in-country representative offices (ROs) or file short-term temporary activities to lawfully operate in mainland China. Without recognized legal status, INGOs are prohibited from working in the country, and even with legal recognition, INGO programming is limited to government-approved activities.

[^ngo-clarification]: "Civil society" is a broad concept that includes a range of organizational forms, both formal and informal. In this paper, we look at nonprofit organizations or third sector organizations---formal non-profit-seeking organizations that are neither privately-owned nor government-controlled. We define international NGOs as organizations that work predominantly in at least one other country outside of their home country.

As an authoritarian state with a growing nonprofit sector, China's regulatory regime for INGOs serves as an example for understanding how autocracies use restrictive laws as policy tools as they engage with international civil society. The policies implemented by China's 2016 ONGO law serve dueling purposes that both harness the benefits and reduce the risks of INGOs. The law is designed to decrease demand for foreign civil society organizations and limit the influence they might have over domestic policies [@DeMattee:2018; @DeMattee:2019] while also allowing the regime to benefit from the expertise, services, and reputational advantages of working partnering with international civil society [@Plantan:2022; @Heiss:2019]. These dual demands lead to differential enforcement and divergence between formal *de jure* regulations and the *de facto* implementation of those laws [@ChaudhryHeiss:2022], as the government enforces regulations selectively depending on regime preferences for INGO services.

In this paper, we use the case of China to explore how the demands to both restrict and embrace INGOs have shaped the international nonprofit sector in the five years since the 2016 ONGO Law came into effect. We examine how the ONGO Law has influenced the population of INGOs that have successfully registered in China, as well as the association between factors such as INGO issue areas and the operational space granted to them in the country. We contribute to a growing body of research on the effects of closing civic space on INGO behavior and strategies. First, we expand work that uses institutional approaches to understanding civil society regulations, which argues that states strategically regulate INGOs to maintain regime stability, choosing forms of regulation and repression that minimize threats to internal political stability [@Chaudhry:2022; @DeMattee:2018; @DeMattee:2019; @Heiss:2019]. Second, we explore how authoritarian civil society regulations act as policy tools that regimes can leverage to shift INGO behavior in their favor and how these laws reshape and tame organizational strategies [@Li:2020; @WuNie:2021; @LianMurdie:2023]. Third, we explore how these regulations lead to differentiated application of the law depending on organizational characteristics. We argue that INGO issue areas, missions, and pre-existing relationships with local government officials influence the degree of operating space available for INGOs.

By focusing on INGO operational space in China, we provide a unique empirical view into the *de facto* regulatory environment for foreign civil society in an authoritarian state. Broader indexes of civil society openness---such as the core civil society index from the Varieties of Democracy project [@LianMurdie:2023; @ChaudhryHeiss:2022]---capture the general regulatory environment for all civil society more broadly, but do not reflect the specific implementation of laws for INGOs. In this paper, we use a mixed methods approach to explore the regulatory environment for INGOs in particular, providing a much richer description and analysis of the impact of INGO laws.

Below, we provide a brief overview of China's 2016 ONGO Law and the different regulatory mechanisms it provides for INGOs to obtain legal access to the country. We then synthesize existing research on authoritarian regulation of INGOs and the differential application of these laws based on regime preferences and organizational characteristics, and propose a set of empirical expectations based on this literature. We test these expectations with a mixed methods research design, combining Bayesian analysis of administrative data from all `r nrow(ongo)` formally registered INGOs with a comparative case study of Greenpeace and The Nature Conservancy, two environmental INGOs that took different approaches to operating in China after 2017. We conclude with a broader discussion of authoritarian civil society regulations and propose future approaches for exploring the practical operating environment for INGOs working in such regimes.

@@

International NGOs began to play an important role in China in the early 1990s as part of the country's Reform and Opening-Up policy, and global civil society has played pivotal roles in providing funding, organizational and program management skills, and catalytic support for China's emerging domestic nonprofit sector [@Yin:2009]. Before the 2016 ONGO Law, the regulation of INGOs in China was ambiguous [@Shieh:2018; @Spires:2022]. The only legal document relevant to INGOs was a section of the Regulation on the Management of Foundations, which required that foreign foundations register with the Ministry of Civil Affairs, the regulating agency for domestic nonprofits. However, the actual application of the regulation was very limited: fewer than 30 foreign foundations registered, compared to estimates of over 7,000 INGOs working in China in 2016 [@YeHuang:2018; @Spires:2020]. Many INGOs gained legal status by registering as foreign-invested companies [@Sidel:2016], allowing them to hire local employees [@HolbigLang:2022; @Yin:2009]. However, this legal form is not quite compatible with the non-profit-distributing, social service orientation of civil society [@Ye:2021]. INGOs registered as foreign-invested companies had to navigate incongruous for-profit regulations, such as rules protecting shareholders' investment returns.

The legal framework governing INGOs was overhauled by the 2016 ONGO Law, which applies to all overseas (including Hong Kong, Macau, and Taiwan) NGOs that conduct not-for-profit activities in mainland China. The ONGO Law provides two pathways for INGOs to lawfully operate in China (see @fig-ro-ta-pathways): (1) submitting an application for formal permanent registration as registered organizations (ROs), or (2) filing documentation of temporary activities with provincial Public Security Bureaus. Formal registration as an RO allows INGOs to maintain an ongoing presence in the country, while temporary activities are intended for INGOs that need to conduct one-off programs that last no longer than one year. Using one of these two channels to obtain legal status became mandatory in 2017. INGOs are authorized to operate within China on a province-by-province basis. Many organizations are allowed to work in all 32 mainland provincial administrative regions, but many are also only authorized to work in a handful, or even only a single province.

```{r fig-ro-ta-pathways, fig.width=6, fig.height=3}
#| echo: false
#| include: true
#| warning: false
#| message: false
#| fig-cap: "Two possible pathways for international NGOs to legally work in 
#|   China under the 2016 ONGO law"

@@

Registering as an RO under the ONGO Law is a complex process that requires the consent of a Chinese Professional Supervisory Unit (PSU)---a governmental or quasi-governmental agency related to an INGO's issue area that oversees and co-supervises the INGO with provincial Public Security Bureaus. Obtaining the consent of a PSU to apply for registration has become a key challenge for INGOs interested in formal registration, as potential PSUs have no legal obligation to assume such a demanding role [@Jia:2017; @Shieh:2018; @YeHuang:2018]. INGOs vary in their ability to attract partner PSUs, and organizations that struggle to find suitable PSUs face limitations in their scope and geographic reach. After successfully gaining legal status, INGOs must follow the law's operational regulations, such as no fundraising within mainland China, only channeling funds through the RO's or PSU's bank accounts, filing reports regularly, and staying within the geographical and issue scopes as approved. INGOs undertaking temporary activities also need to find an eligible Chinese partner---including governmental, quasi-governmental, or non-profit organizations---and must keep their programs within approved limits, but the process is less onerous.
